Static contract checking for Haskell

### 2011-12-21 paper/paper.tex
Paper: updating phi_cf section per experiments.

  We also use projections in some versions of the case-expression
  translation, to Skolemize, but this is just a happy coincidence.  In
  particular, user programs will never contain projections.  If a user
  wants to project then they need to write there own projection
  function using pattern matching.
\item $(\phi_\text{disjoint})$ Distinct term constructors have
  disjoint ranges.  Note that we don't axiomitize disjointness of term
  constructors from distinct \emph{data types}.
\item $(\phi_\text{cf})$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\gets$: if we pack crash-free values in a data constructor,
    the resulting value is crash-free.  If $K$ has arity $n$
    then we assert $\ctrans{K \in \cfc^n \to \cfc}^-$.

  \item $\to$: if a value $t$ of type $T$\footnote{
    ``type'' is used loosely here: it means the constructor 
      head is from the type $T$.
    } is crash-free then
    every value packed in it is crash-free, since we can use
    a case expression to extract any packed value.
    \note[nc]{XXX: This translation is ad-hoc.}

  Note that this is not true for arbitrary functions: a function is not required
  to use all of its arguments. For example, \texttt{fst} is crash-free if and only
  if the first argument of the pair is crash-free. The second argument
  is irrelevant.

@@

\begin{figure*}
  $$ \ruleform{\ttrans{data~def} = \Phi} $$
$$\ttrans{\mbox{data } T = K_1 ~a_1, \dots, K_n ~a_n} = \phi_\text{injective} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
 \hspace{5.6cm}where
  \begin{center}
    \newcommand{\fullkxs}[1]{\full{K}_#1(\vec{x})}
    \begin{align*}
      \phi_\text{injective} &= \bigwedge_{1 \leq i \leq n}~\left( \forall \vec{x}.~\min{\fullkxs i} \to \bigwedge_{1 \leq j \leq a_i} x_j = \proj{j}{K_i}(\fullkxs i) \right) \\
      \phi_\text{disjoint} &= \!\!\!\bigwedge_{1 \leq i < j \leq n}\!\!\!~\left(\forall \vec{x},\vec{y}.~ (\min{\fullkxs i} \lor \min{\fullkxs j}) \to \fullkxs {i} \neq \fullkxs {j} \right)\\
      \phi_\text{cf} &= 
        \bigwedge_{1 \leq i \leq n}~\left( \ctrans{K_i \in \cfc^{a_i} \to \cfc}^- 
          \land \forall \vec{x}.~\min{\fullkxs i}
                  \land \cf{\fullkxs i} \to \bigwedge_{1 \leq j \leq a_i} \cf{x_j} \right) \\
      \phi_\text{lazy} &= \bigwedge_{1 \leq i \leq n}~\left( \forall \vec{x}.~(\min{\fullkxs i} \to \fullkxs {i} \neq \unr \land \fullkxs {i} \neq \bad \right)
    \end{align*}
  \end{center}
  \caption{$\ttrans{}$ -- Data type translation}
  \label{ktrans}
\end{figure*}

>>>>

  We also use projections in some versions of the case-expression
  translation, to Skolemize, but this is just a happy coincidence.  In
  particular, user programs will never contain projections.  If a user
  wants to project then they need to write there own projection
  function using pattern matching.
\item $(\phi_\text{disjoint})$ Distinct term constructors have
  disjoint ranges.  Note that we don't axiomitize disjointness of term
  constructors from distinct \emph{data types}.
\item $(\phi_\text{cf})$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\gets$: if we pack crash-free values in a data constructor,
    the resulting value is crash-free.  If $K$ has arity $n$ then we
    assert $\ctrans{K \in \cfc^n \to \cfc}^-$.  I.e., we simply treat
    K as a strict \cfc function.

  \item $\to$: if a value $t$ of type $T$\footnote{ ``type'' is used
      loosely here: it means the constructor head is from the type
      $T$.  } is crash-free then every value packed in it is
    crash-free, since we can use a case expression to extract any
    packed value.  More precisely, we assert
    \[
    \forall \vec{x}. \min{K~\vec{x}} \land \cf{K~\vec{x}}
    \to \bigwedge_{1 \le i \le n} \cf{x_i}.
    \]
    Unlike the $\gets$-direction of the translation, where we use
    $\ctrans{}$ (principled), this direction of the translation is
    ad-hoc.  Interestingly, dropping the $\min$ here causes
    problems. The \texttt{lem\_eqNat\_trans} in
    \texttt{yes/add-symmetric.hs} becomes unprovable (or at least,
    really slow), and Paradox is no longer able to find small
    counter-models to \texttt{lem\_add\_symmetric} when
    \texttt{lem\_eqNat\_trans}'s $\cfc^3 \to \cfc$ contract is removed.

  Note that $\phi_\text{cf}$ is not true for arbitrary functions: a function is not required
  to use all of its arguments. For example, \texttt{fst} is crash-free if and only
  if the first argument of the pair is crash-free. The second argument
  is irrelevant.

@@

\begin{figure*}
  $$ \ruleform{\ttrans{data~def} = \Phi} $$
$$\ttrans{\mbox{data } T = K_1 ~a_1, \dots, K_n ~a_n} = \phi_\text{injective} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
 \hspace{5.6cm}where
  \begin{center}
    \newcommand{\fullkxs}[1]{\full{K}_#1~\vec{x}}
    \begin{align*}
      \phi_\text{injective} &= \bigwedge_{1 \leq i \leq n}~\left( \forall \vec{x}.~\min{\fullkxs i} \to \bigwedge_{1 \leq j \leq a_i} x_j = \proj{j}{K_i}(\fullkxs i) \right) \\
      \phi_\text{disjoint} &= \!\!\!\bigwedge_{1 \leq i < j \leq n}\!\!\!~\left(\forall \vec{x},\vec{y}.~ (\min{\fullkxs i} \lor \min{\fullkxs j}) \to \fullkxs {i} \neq \fullkxs {j} \right)\\
      \phi_\text{cf} &= 
        \bigwedge_{1 \leq i \leq n}~\left( \ctrans{K_i \in \cfc^{a_i} \to \cfc}^- 
          \land \forall \vec{x}.~\min{\fullkxs i}
                  \land \cf{\fullkxs i} \to \bigwedge_{1 \leq j \leq a_i} \cf{x_j} \right) \\
      \phi_\text{lazy} &= \bigwedge_{1 \leq i \leq n}~\left( \forall \vec{x}.~(\min{\fullkxs i} \to \fullkxs {i} \neq \unr \land \fullkxs {i} \neq \bad \right)
    \end{align*}
  \end{center}
  \caption{$\ttrans{}$ -- Data type translation}
  \label{ktrans}
\end{figure*}
